---
title: "Dengue Epidemiology Dashboard: Sri Lanka"
format:
  dashboard:
    theme: cosmo
    orientation: rows
    scrolling: true
    nav-buttons:
      - icon: github
        href: https://github.com/thiyangt/denguedatahub
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

# ---------------------------------------------------------------------------

# install.packages(c("denguedatahub","tidyverse","lubridate","tsibble",
#                     "feasts","fabletools","fable","timetk","anomalize",
#                     "plotly","dygraphs","xts","leaflet","sf","DT","viridis",
#                     "crosstalk"))
# devtools::install_github("thiyangt/ceylon") 
# ---------------------------------------------------------------------------

library(denguedatahub)
library(tidyverse)
library(lubridate)
library(tsibble)
library(feasts)
library(fabletools)
library(timetk)
library(plotly)
library(dygraphs)
library(xts)
library(leaflet)
library(sf)
library(DT)
library(viridis)

# Accent colour used throughout
accent <- "#c0392b"

# --- Base data ------------------------------------------------------------
data(srilanka_weekly_data)

sl <- srilanka_weekly_data |>
  mutate(start.date = mdy(start.date)) |>
  arrange(start.date)

# --- National weekly series ----------------------------------------------
national <- sl |>
  group_by(year, week, start.date) |>
  summarise(cases = sum(cases, na.rm = TRUE), .groups = "drop") |>
  arrange(start.date)

# Regular weekly tsibble for STL / decomposition
national_tsbl <- sl |>
  mutate(year_week = yearweek(start.date)) |>
  group_by(year_week) |>
  summarise(cases = sum(cases, na.rm = TRUE), .groups = "drop") |>
  as_tsibble(index = year_week) |>
  fill_gaps(cases = 0L)

# --- Annual totals --------------------------------------------------------
annual <- sl |>
  group_by(year) |>
  summarise(cases = sum(cases, na.rm = TRUE), .groups = "drop") |>
  arrange(year)

# --- District totals ------------------------------------------------------
district_tot <- sl |>
  group_by(district) |>
  summarise(total_cases  = sum(cases, na.rm = TRUE),
            peak_weekly   = max(cases, na.rm = TRUE),
            .groups = "drop") |>
  arrange(desc(total_cases)) |>
  mutate(key = toupper(gsub("[^A-Za-z]", "", district)))

# --- Headline numbers for value boxes ------------------------------------
total_cases     <- sum(sl$cases, na.rm = TRUE)
worst_year      <- annual$year[which.max(annual$cases)]
worst_year_n    <- max(annual$cases)
worst_district  <- district_tot$district[1]
latest_row      <- national |> slice_max(start.date, n = 1)
latest_week_lab <- format(latest_row$start.date, "%d %b %Y")
latest_week_n   <- latest_row$cases
```

```{=html}
<style>
/* Force readable text + icons on the coloured KPI cards */
.bslib-value-box .value-box-title,
.bslib-value-box .value-box-value { color: #ffffff !important; }
.bslib-value-box .value-box-area  { color: #ffffff !important; }
.bslib-value-box .value-box-showcase,
.bslib-value-box .value-box-showcase .bi { color: #ffffff !important; opacity: .95; }
.bslib-value-box .value-box-value { font-weight: 700; }

/* Darker top navigation bar */
.navbar,
.quarto-dashboard .navbar {
  background-color: #14293a !important;
  border-bottom: 1px solid rgba(255, 255, 255, 0.08) !important;
}
.navbar .navbar-brand,
.navbar .navbar-title,
.quarto-dashboard .navbar .navbar-title {
  color: #ffffff !important;
  font-weight: 600;
}
.navbar .nav-link,
.navbar .navbar-nav .nav-link,
.quarto-dashboard .navbar .nav-link {
  color: rgba(255, 255, 255, 0.78) !important;
}
.navbar .nav-link:hover {
  color: #ffffff !important;
  background-color: rgba(255, 255, 255, 0.08) !important;
  border-radius: 6px;
}
.navbar .nav-link.active,
.navbar .navbar-nav .nav-link.active,
.quarto-dashboard .navbar .navbar-nav .nav-link.active,
.navbar .nav-item.active .nav-link {
  color: #ffffff !important;
  background-color: rgba(255, 255, 255, 0.14) !important;
  border-radius: 6px;
  box-shadow: inset 0 -3px 0 #c0392b;
}
/* nav-buttons / tool icons (e.g. GitHub) */
.navbar .quarto-navbar-tools a,
.navbar .navbar-nav .nav-link .bi,
.quarto-dashboard .navbar .quarto-dashboard-links a {
  color: rgba(255, 255, 255, 0.9) !important;
}
/* Tidy crosstalk checkbox lists (district filters): box inline with label */
.crosstalk-input .checkbox { margin: 2px 0; }
.crosstalk-input .checkbox label {
  display: flex;
  align-items: center;
  gap: 6px;
  padding-left: 0;
  font-weight: 400;
  font-size: 0.85rem;
  line-height: 1.2;
}
.crosstalk-input .checkbox input[type="checkbox"] {
  position: static;
  margin: 0;
  flex: 0 0 auto;
}
</style>
```

# Overview

## Row {height="175px"}

```{r}
#| content: valuebox
list(title = "Total reported cases · 2006 to present",
     value = format(total_cases, big.mark = ","),
     icon  = "virus2",
     color = "#2f6690")
```

```{r}
#| content: valuebox
list(title = paste0("Cases in the worst year · ", worst_year),
     value = format(worst_year_n, big.mark = ","),
     icon  = "graph-up-arrow",
     color = "#a4243b")
```

```{r}
#| content: valuebox
list(title = "Highest-burden district",
     value = worst_district,
     icon  = "pin-map-fill",
     color = "#1d7874")
```

```{r}
#| content: valuebox
list(title = paste0("Latest reported week · ", latest_week_lab),
     value = format(latest_week_n, big.mark = ","),
     icon  = "calendar3-week",
     color = "#3d4f7c")
```

## Row {height="600px"}

### Column {width=65%}

```{r}
#| title: "National weekly dengue cases (drag the range selector to zoom)"
nat_xts <- xts(national$cases, order.by = national$start.date)
dygraph(nat_xts, width = "100%", height = 450) |>
  dyOptions(colors = accent, fillGraph = TRUE, strokeWidth = 1.3) |>
  dyHighlight(highlightCircleSize = 4) |>
  dyRangeSelector(height = 30) |>
  dyAxis("y", label = "Reported cases")
```

### Column {width=35%}

```{r}
#| title: "Annual Dengue Cases"
plot_ly(annual, x = ~year, y = ~cases, type = "bar",
        marker = list(
          color = ~cases,
          # Yellow (fewest cases) -> orange -> dark red (most cases)
          colorscale = list(c(0,    "#ffffb2"),
                            c(0.25, "#fecc5c"),
                            c(0.5,  "#fd8d3c"),
                            c(0.75, "#f03b20"),
                            c(1,    "#bd0026")),
          reversescale = FALSE,
          line = list(color = "white", width = 0.5)),
        hovertemplate = "Year %{x}<br>%{y:,} cases<extra></extra>") |>
  layout(xaxis = list(title = ""), yaxis = list(title = "Cases"),
         margin = list(t = 10))
```

# Districts

## Row {height="560px"}

### Column {width=55%}

```{r}
#| title: "Cumulative reported cases by district"
# District boundaries from the ceylon package (SLD99 grid -> WGS84 for leaflet)
library(ceylon)
data(district)

sl_dist <- district |>
  filter(!is.na(DISTRICT), DISTRICT != "[UNKNOWN]") |>
  mutate(key = toupper(gsub("[^A-Za-z]", "", DISTRICT))) |>
  st_transform(4326) |>
  left_join(district_tot, by = "key")

pal <- colorNumeric("YlOrRd", domain = sl_dist$total_cases)

leaflet(sl_dist) |>
  addProviderTiles("CartoDB.Positron") |>
  addPolygons(
    fillColor = ~pal(total_cases), weight = 1, color = "white",
    fillOpacity = 0.85,
    highlightOptions = highlightOptions(weight = 2.5, color = "#333",
                                        fillOpacity = 0.95, bringToFront = TRUE),
    label = ~sprintf("%s: %s cases", DISTRICT,
                     format(total_cases, big.mark = ","))
  ) |>
  addLegend(pal = pal, values = ~total_cases, position = "bottomright",
            title = "Total cases<br>2006–present", opacity = 0.9)
```

### Column {width=45%}

```{r}
#| title: "District ranking"
district_tot |>
  transmute(District = district,
            `Total cases` = total_cases,
            `Peak weekly` = peak_weekly) |>
  datatable(options = list(pageLength = 10, dom = "tip"),
            rownames = FALSE) |>
  formatCurrency(c("Total cases", "Peak weekly"), currency = "", digits = 0)
```

## Row {height="620px"}

### Column {width=75%}

```{r}
#| title: "Weekly cases by district (tick districts on the right to overlay them)"
library(crosstalk)

# Shared data links the plot and the checkbox list, even in separate columns
sd_dist <- SharedData$new(sl)

plot_ly(sd_dist, x = ~start.date, y = ~cases, color = ~district,
        colors = viridis(length(unique(sl$district)), option = "turbo"),
        type = "scatter", mode = "lines",
        hovertemplate = "%{x|%d %b %Y}<br>%{y} cases<extra>%{fullData.name}</extra>") |>
  layout(xaxis = list(title = ""), yaxis = list(title = "Cases"),
         margin = list(t = 10, b = 30), showlegend = FALSE)
```

### Column {width=25%}

```{r}
#| title: "Districts"
htmltools::tags$div(
  htmltools::tags$p(
    htmltools::HTML("Tick one or more districts to overlay their series; "),
    htmltools::HTML("leave <b>all unticked</b> for every district."),
    style = "color:#6c757d; font-size:0.8rem; margin:0 0 8px 0; line-height:1.4;"
  ),
  htmltools::tags$div(
    filter_checkbox("district_pick", "", sd_dist, ~district),
    style = "max-height:520px; overflow-y:auto; padding-right:4px;"
  )
)
```

# By Year

## Row {height="700px"}

```{r}
#| title: "Total cases by district and year"
dist_order <- sl |>
  group_by(district) |>
  summarise(t = sum(cases, na.rm = TRUE), .groups = "drop") |>
  arrange(t)

dist_year_hm <- sl |>
  group_by(district, year) |>
  summarise(cases = sum(cases, na.rm = TRUE), .groups = "drop") |>
  mutate(district = factor(district, levels = dist_order$district),
         cases = as.integer(round(cases)),
         tip = paste0(district, " \u00b7 ", year, "<br>",
                      formatC(cases, format = "d", big.mark = ","), " cases"))

p_dy <- ggplot(dist_year_hm,
               aes(factor(year), district, fill = cases, text = tip)) +
  geom_tile(colour = "white", linewidth = 0.2) +
  scale_fill_viridis_c(option = "C", trans = "sqrt", name = "Cases") +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggplotly(p_dy, tooltip = "text")
```

## Row {height="720px"}

```{r}
#| title: "District case maps, year by year"
#| fig-width: 14
#| fig-height: 7
library(ceylon)
data(district)

dist_year_map <- sl |>
  mutate(key = toupper(gsub("[^A-Za-z]", "", district))) |>
  group_by(year, key) |>
  summarise(cases = sum(cases, na.rm = TRUE), .groups = "drop")

sl_poly <- district |>
  filter(!is.na(DISTRICT), DISTRICT != "[UNKNOWN]") |>
  mutate(key = toupper(gsub("[^A-Za-z]", "", DISTRICT))) |>
  st_transform(4326) |>
  st_simplify(dTolerance = 0.005, preserveTopology = TRUE) |>
  dplyr::select(key)

map_year <- sl_poly |>
  dplyr::left_join(dist_year_map, by = "key") |>
  dplyr::filter(!is.na(year))                 # drop the stray "NA" panel

ggplot(map_year) +
  geom_sf(aes(fill = cases), colour = "white", linewidth = 0.08) +
  facet_wrap(~ year, ncol = 7) +
  scale_fill_viridis_c(option = "C", trans = "sqrt", name = "Cases",
                       labels = scales::label_comma(),
                       guide = guide_colorbar(title.position = "top",
                                              title.hjust = 0.5)) +
  labs(x = NULL, y = NULL) +
  theme_void(base_size = 10) +
  theme(
    strip.text        = element_text(face = "bold", size = 10,
                                     margin = margin(b = 3)),
    panel.spacing     = unit(3, "pt"),
    legend.position   = "bottom",
    legend.key.width  = unit(40, "pt"),
    legend.key.height = unit(9, "pt"),
    legend.title      = element_text(size = 10),
    legend.text       = element_text(size = 8),
    plot.margin       = margin(6, 10, 6, 10)
  )
```

## Row {height="520px"}

### Column {width=75%}

```{r}
#| title: "Annual total cases by district (tick districts on the right to overlay them)"
library(crosstalk)

annual_dist <- sl |>
  group_by(district, year) |>
  summarise(cases = sum(cases, na.rm = TRUE), .groups = "drop")

# Shared data links the plot and the checkbox list, even in separate columns
sd_annual <- SharedData$new(annual_dist)

plot_ly(sd_annual, x = ~year, y = ~cases, color = ~district,
        colors = viridis(length(unique(annual_dist$district)), option = "turbo"),
        type = "scatter", mode = "lines",
        hovertemplate = "%{x}<br>%{y:,} cases<extra>%{fullData.name}</extra>") |>
  layout(xaxis = list(title = ""), yaxis = list(title = "Total cases"),
         margin = list(t = 10), showlegend = FALSE)
```

### Column {width=25%}

```{r}
#| title: "Districts"
htmltools::tags$div(
  htmltools::tags$p(
    htmltools::HTML("Tick one or more districts to overlay their annual totals; "),
    htmltools::HTML("leave <b>all unticked</b> for every district."),
    style = "color:#6c757d; font-size:0.8rem; margin:0 0 8px 0; line-height:1.4;"
  ),
  htmltools::tags$div(
    filter_checkbox("district_pick_yr", "", sd_annual, ~district),
    style = "max-height:380px; overflow-y:auto; padding-right:4px;"
  )
)
```

# Seasonality & Trend

## Row {height="200px"}

### Column {width=50%}

```{r}
#| title: "STL — observed cases"
stl_df <- national_tsbl |>
  model(STL(cases ~ season(period = 52))) |>
  components() |>
  as_tibble() |>
  mutate(date = as.Date(year_week))

plot_ly(stl_df, x = ~date, y = ~cases, type = "scatter", mode = "lines",
        line = list(color = "#1b3a4b")) |>
  layout(xaxis = list(title = ""), yaxis = list(title = "Cases"),
         margin = list(t = 10))
```

### Column {width=50%}

```{r}
#| title: "STL — trend"
plot_ly(stl_df, x = ~date, y = ~trend, type = "scatter", mode = "lines",
        line = list(color = "#2f6690")) |>
  layout(xaxis = list(title = ""), yaxis = list(title = "Trend"),
         margin = list(t = 10))
```

## Row {height="200px"}

### Column {width=50%}

```{r}
#| title: "STL — seasonal (annual, period = 52)"
plot_ly(stl_df, x = ~date, y = ~season_52, type = "scatter", mode = "lines",
        line = list(color = "#1d7874")) |>
  layout(xaxis = list(title = ""), yaxis = list(title = "Seasonal"),
         margin = list(t = 10))
```

### Column {width=50%}

```{r}
#| title: "STL — remainder"
plot_ly(stl_df, x = ~date, y = ~remainder, type = "scatter", mode = "lines",
        line = list(color = "#8a8d91")) |>
  layout(xaxis = list(title = ""), yaxis = list(title = "Remainder"),
         margin = list(t = 10))
```


## Row {height="700px"}

```{r}
#| title: "Weekly dengue Cases Heatmap"
hm <- sl |>
  group_by(Year = year(start.date), Week = isoweek(start.date)) |>
  summarise(cases = sum(cases, na.rm = TRUE), .groups = "drop")

p_hm <- ggplot(hm, aes(Week, factor(Year), fill = cases)) +
  geom_tile(colour = "white", linewidth = 0.2) +
  scale_fill_viridis_c(option = "C", name = "Cases") +
  labs(x = "ISO week", y = NULL) +
  theme_minimal(base_size = 12)
ggplotly(p_hm)
```

# Outbreaks

## Row {height="690px"}

### Column {width=64%}

```{r}
#| title: "Anomaly detection"
library(anomalize)
national |>
  time_decompose(cases, method = "stl", frequency = 52, trend = 104) |>
  anomalize(remainder, method = "iqr", alpha = 0.05) |>
  time_recompose() |>
  plot_anomalies(time_recomposed = TRUE, ncol = 1, alpha_dots = 0.4) +
  labs(x = NULL, y = "Cases") +
  theme_minimal(base_size = 11)
```

### Column {width=36%}

```{r}
#| title: "Detected outbreak weeks"
anom <- national |>
  time_decompose(cases, method = "stl", frequency = 52) |>
  anomalize(remainder, method = "iqr", alpha = 0.05) |>
  time_recompose()

anom |>
  filter(anomaly == "Yes") |>
  as_tibble() |>
  transmute(Week  = format(start.date, "%Y-%m-%d"),
            Cases = observed) |>
  arrange(desc(Cases)) |>
  datatable(options = list(pageLength = 12, dom = "tip"), rownames = FALSE)
```

# Data

## Row {height="170px"}

```{r}
#| content: valuebox
list(title = "Weekly observations",
     value = format(nrow(srilanka_weekly_data), big.mark = ","),
     icon  = "table",
     color = "#2f6690")
```

```{r}
#| content: valuebox
list(title = "Districts reported",
     value = dplyr::n_distinct(sl$district),
     icon  = "diagram-3-fill",
     color = "#1d7874")
```

```{r}
#| content: valuebox
list(title = "Reporting period",
     value = paste0(format(min(sl$start.date), "%Y"), "–",
                    format(max(sl$start.date), "%Y")),
     icon  = "calendar-range",
     color = "#3d4f7c")
```

```{r}
#| content: valuebox
list(title = "Data source",
     value = "Epi Unit, MoH",
     icon  = "hospital",
     color = "#a4243b")
```

## Row

### Column {width=72%}

```{r}
#| title: "Browse the raw weekly data — search or filter any column"
srilanka_weekly_data |>
  transmute(
    Year         = factor(year, levels = sort(unique(year))),
    Week         = factor(week, levels = sort(unique(week))),
    `Start date` = start.date,
    `End date`   = end.date,
    District     = district,
    Cases        = as.integer(cases)
  ) |>
  datatable(filter = "top", rownames = FALSE, width = "100%",
            options = list(pageLength = 12, dom = "tip", scrollX = TRUE,
                           autoWidth = TRUE,
                           order = list(list(0, "desc"), list(1, "desc"))))
```

### Column {width=28%}

::: {.card title="What the data contains"}

`srilanka_weekly_data` from the **denguedatahub** package — one row per district per
epidemiological week, sourced from the Epidemiology Unit, Ministry of Health, Sri Lanka
(Weekly Epidemiological Reports).

| Column | Meaning |
|---|---|
| `year` | Epidemiological year |
| `week` | Epi week number (1–52/53) |
| `start.date` | First day of the reporting week |
| `end.date` | Last day of the reporting week |
| `district` | Administrative reporting area |
| `cases` | Notified suspected dengue cases |

:::

::: {.card title="Data sources & citations"}

**Dengue case data**

Talagala T (2026). *denguedatahub: A Tidy Format Datasets of Dengue by Country.* R package version 4.1.0. doi:10.32614/CRAN.package.denguedatahub. <https://CRAN.R-project.org/package=denguedatahub>

Original source: Epidemiology Unit, Ministry of Health, Sri Lanka — Weekly Epidemiological Reports. <https://www.epid.gov.lk/>

**District boundaries**

Talagala T (2026). *ceylon: Sri Lanka Maps.* R package version 1.1.0. <https://github.com/thiyangt/ceylon>


:::
